---
title: "Statistical Predictive Model for Automobile Pricing"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

First we ensure that the required packages for our analysis are installed.
```{r}
#install.packages("tidyverse")
library(tidyr)
library(readr)
library(dplyr)
library(ggplot2)
source("ford.R", local = TRUE)
```

Now we load the data. The data load uses the `read_csv` function from the `dplyr` package.
```{R}
data <- read_csv("craigslistVehicles_semi.csv");
```


Lets quickly summarise the data that was loaded
```{R}
summary(data);
#plot(as.numeric(data$year),as.numeric(data$odometer))
#levels(data$type)
#summary(data$type)
#data %>%group_by(type)%>%summarise(count = n()) %>% View
```


We can see from the summary that we have observations in our dataset that are incomplete. 
We have `NA`s for several variables, and we shall remove these incomplete observations.
``` {R}

#complete <- data[complete.cases(data),]
complete <- data %>%  
    drop_na(odometer) %>%
    drop_na(year) %>%
    drop_na(make) %>%
    drop_na(manufacturer)
```

Now that we have done a bit more clean up, lets get some summary stats for the manufacturers
```{R}

# clean manufacturers

complete %>% group_by(manufacturer)%>%summarise(count = n())
sig_manu <- complete %>% group_by(manufacturer)%>%summarise(count = n()) %>% filter(count>1000)

complete <- complete %>% filter(manufacturer %in% sig_manu$manufacturer)
#complete %>% group_by(manufacturer)%>%summarise(count = n()) %>% View

# clean price

#complete$price <-with(complete,as.numeric(as.character(price)))
quantile(complete$price,probs = seq(0,1,.01))
quantile(complete$price,.05)
summary(complete$price)
price_dist<-complete %>%filter(price< quantile(complete$price,.99)) %>% group_by(price) %>% summarise(count = n()) 
```
Lets plot the distribution of the prices
```{R}
ggplot(data=price_dist,aes(x=price,y=count)) + geom_point()
```
We can still see some of the extreme observations on te lower end. Specifically, there are many observations that have a `price` of `0`. These typically mean that the seller is trying to negotiate the best offer from potential buyers. We can see that upto the 10th percentile, the `price` is `0`. We are excluding these observations, as the price is not indicative of the actual expected sale price.

```{r}
price_dist<-complete %>%filter(price > quantile(complete$price,.1) & price < quantile(complete$price,.95)) %>% group_by(price) %>% summarise(count = n()) 
ggplot(data=price_dist,aes(x=price,y=count)) + geom_point()
```

## Including Plots

You can also embed plots, for example:

```{r price_dist, echo=FALSE}
ggplot(data=price_dist,aes(x=price,y=count)) + geom_point()
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

[ford]["https://www.ford.com"]
